---
title: "Assignment 1 - 02429 Analysis of correlated data: Linear mixed models"
subtitle: "Group 4"
date: "Monday September 28, 2026"
author: "Lachlan Casey, Boglárka Dorottya Kováts, Rune Ringgaard Jørgensen"
format:
  pdf:
    geometry:
     - left=2cm
     - right=2cm
     - bottom=3cm
    fig-pos: h
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{fancyhdr}
        \pagestyle{fancy}
        \fancyhf{}
        \fancyhead[C]{Assignment 1 - Group 4}
        \fancyfoot[C]{\thepage}
execute: 
  echo: false
bibliography: references.bib
---

```{r}
#| label: load data 1
#| include: false


source("R/exercise_1.R")

```

# Exercise 1

## Visualisation

Exploring the data, @fig-boxplot-bulls shows the percentage of conception by bull, displayed in a box plot. @fig-boxplot-bulls shows a large variance in the samples from bull 4, while others, e.g., bull 3 have a lower variance. Furthermore, bull 5 has a high percentage of conception compared to the rest of the sample.

```{r}
#| label: fig-boxplot-bulls
#| echo: false
#| fig-cap: Percentage of conception of bulls.
#| fig-asp: .8
#| fig-pos: h

par(mai=c(1, 1, .25, 0.5))
boxplot(perc ~ bull, data = df, xlab = 'Bull', ylab = '% of Conceptions', las=1, col=2:7)
par(mai=c(1, 1, 1, 1))

```

## Estimation of percentage of conception

Considering the bulls as a random sample of a population, and assuming that observations from each bull are correlated, the percentage of conception can be estimated with a random-effects model with bull as a random factor.

### Model

$$
y_{ij} = \mu + a_{i} + \epsilon_{ij}
$$ {#eq-model1}

Where:

- $y_{ij}$ represents the $j$-th percentage measurement for the $i$-th bull.

- $\mu$ is the overall true average percentage of conceptions across the entire population.

- $a_{i}$ is the random effect for the $i$-th bull, assumed to be normally distributed with a mean of zero and a variance of $\sigma_{a}^{2}$ ($a_{i} \sim N\left( 0,\sigma_{a}^{2} \right)$).

- $\epsilon_{ij}$ is the residual measurement error for each specific sample, assumed to be normally distributed with a mean of zero and a variance of $\sigma_{\epsilon}^{2}$ ($\epsilon_{ij} \sim N\left( 0,\sigma_{\epsilon}^{2} \right)$).

 

## Results

$\mu=$ **`r estimate`**

Bull-to-bull variability (${\widehat{\sigma}}_{a}^{2}$): `r round(sigma_bull, 2)`

Within-bull/Residual variability (${\widehat{\sigma}}_{\epsilon}^{2}$): `r round(sigma_within, 2)`

To calculate the heritability parameter, $h$, @eq-heritability is used.

$$
h = \frac{4\sigma_{a}^{2}}{\sigma_{a}^{2} + \sigma_{\epsilon}^{2}}
$$ {#eq-heritability}

$h$ = `r round(heritability * 100, 2)`%

::: {}
:::

# Exercise 2

```{r}
#| label: load data 2
#| include: false


source("R/exercise_2.R")

```

## Visualisation

@fig-rat-plots shows boxplots of breaking strength by site (@fig-site), and by treatment (@fig-treat). @fig-treat could indicate that the treatment does not increase breaking strength, or even decreases it.

::: {#fig-rat-plots layout="[[48,-4,48]]" fig-pos="H"}
```{r}
#| label: fig-site
#| echo: false
#| fig-asp: 1.2
#| fig-cap: Strength by Site

# Set up a 1x2 grid for side-by-side boxplots
# par(mfrow = c(1, 2))

# Boxplot: Raw log(Strength) across the 5 Sites
boxplot(logY ~ site, data = rats_data,
        xlab = 'Site on Back', ylab = 'log(Breaking Strength)',
        col = rainbow(5))
points(1:5, means_site, pch = 23, bg = "black") # Add diamond markers for the means
```

```{r}
#| label: fig-treat
#| echo: false
#| fig-asp: 1.2
#| fig-cap: Strength by Treatment
# Boxplot: Raw log(Strength) across the 2 Treatments
boxplot(logY ~ treat, data = rats_data,
        xlab = 'Treatment', ylab = 'log(Breaking Strength)',
        col = c("lightblue", "lightgreen"))
points(1:2, means_treat, pch = 23, bg = "black")

# Reset the plot grid to 1x1
# par(mfrow = c(1, 1))

```

Visualisations of rat data
:::

## Task i)

An initial linear mixed model, @eq-model2a-lmm, is fitted with interactions between Treatment and Site.

$$
\log Y_{ijk} = \mu + \alpha_{i} + \beta_{j} + (\alpha\beta)_{ij} + d_{k} + \epsilon_{ijk}
$$ {#eq-model2a-lmm}

Where:

- $\log Y_{ijk}$ is the log-transformed breaking energy for the $k$-th rat at the $i$-th site under the $j$-th treatment.

- $\mu$ is the overall average $logY$.

- $\alpha_{i}$ is the fixed main effect of the $i$-th site (1 through 5).

- $\beta_{j}$ is the fixed main effect of the $j$-th treatment (Hyperbaric O2 or Control).

- $(\alpha\beta)_{ij}$ is the fixed interaction effect between the $i$-th site and the $j$-th treatment, representing how the treatment effect might change depending on the specific site.

- $d_{k}$ is the random effect of the $k$-th rat, assumed to be normally distributed with a mean of zero and a variance of $\sigma_{rat}^{2}$ ($d_{k} \sim N\left( 0,\sigma_{rat}^{2} \right)$).

- $\epsilon_{ijk}$ is the residual measurement error, assumed to be normally distributed with a mean of zero and a variance of $\sigma^{2}$ ($\epsilon_{ijk} \sim N\left( 0,\sigma^{2} \right)$).

Running an F-test (`drop1`) shows a p-value value of 0.4292 for the interaction $(\alpha\beta)_{ij}$, i.e., the interaction is insignificant indicating it can be dropped.

A final linear mixed model, @eq-model2b-lmm, is fitted, omitting the interaction between Treatment and Site and keeping rat as a random effect.

$$
\log Y_{ijk} = \mu + \alpha_{i} + \beta_{j} + d_{k} + \epsilon_{ijk}
$$ {#eq-model2b-lmm}

Analysis of variance (ANOVA) of the final model showed that *site* was statistically significant with $p =$ `r round(lmm2_anova[1,6], 4)` and is likely to effect the logarithm of the response $Y$, unlike *treatment* with $p =$ `r round(lmm2_anova[2,6], 4)`, see @tbl-rat-anova.

```{r}
#| tbl-cap: Analysis of variance (ANOVA) of @eq-model2b-lmm.
#| label: tbl-rat-anova

library(knitr)

lmm2_anova %>% 
  rename('p-value' = `Pr(>F)`) %>% 
  kable()

```

The variance from rat to rat is `r round(vars[1, "vcov"], 4)` and the variance of the residual measurement is `r round(vars[2, "vcov"], 4)`. The effect of each site can be seen in @tbl-rat-coef, relative to the intercept which is site 1. Sites 2,3,4 increase the logarithm of the response where site 5 decreases the logarithm of the response. The treatment which is deemed insignificant also decreases the logarithmic response.

```{r}
#| tbl-cap: Fixed effects coefficients of @eq-model2b-lmm.
#| label: tbl-rat-coef

m2_coef %>% 
  rename('p-value' = last_col()) %>% 
  kable()

```

## Task ii)

Site 5 exhibited the lowest estimated marginal mean, as seen in @fig-emm.

```{r}
#| label: fig-emm
#| fig-cap: Estimated Marginal Mean of logY
#| fig-pos: h

plot(site_means, comparisons = TRUE,
     xlab = "EMM",
     ylab = "Site on Back") +
  theme(plot.title = element_text(hjust = 0.5))

```

```{r}
#| label: fig-tukey
#| fig-cap: "95% family-wise confidence level: Site Differences"
#| fig-pos: H
#| fig-asp: .6

par(mai=c(1, 1, .25, 0.5))
plot(mult_site, col=2:11,
     main = "")
par(mai=c(1, 1, 1, 1))

```

As can be seen in @fig-tukey, the pairwise comparisons involving site 5 showed that the 95% confidence intervals for the comparisons between site 5 and sites 2, 3, and 4 did not include zero, indicating significant differences. In contrast, the confidence interval for the comparison between sites 5 and 1 included zero, indicating no statistically significant difference between these two sites.

## Task iii)

Fitting a linear model (@eq-rat-lm) suggest the *treatment* is significant due to the linear model underestimating the "true" variance which the LMM takes into account as can be seen in @tbl-lmfit-coef.

$$
\log(Y) = \mu + \text{site} + \text{treat} + \varepsilon
$$ {#eq-rat-lm}

None of the p-values indicate any statistical significance in @tbl-lmfit-coef.

```{r}
#| label: tbl-lmfit-coef
#| tbl-cap: "Coefficients of a fitted linear model."

lmcoef <- lm_fit_sum$coefficients %>% 
  as.data.frame()

lmcoef <- lmcoef[1:5,]

lmcoef %>% 
  rename("p-value" = last_col()) %>% 
  kable()

```

F-test of the linear model, @eq-rat-lm, shows that reducing the model by removing *site* is possible ($p$ = `r signif(var_p[2,1], 3)`) and that *treatment* is significant ($p$ = `r signif(var_p[3,1], 3)`).

Conversely, as shown previously in the analysis of variance of the LMM model, @tbl-rat-anova, *site* is a statistically significant variable ($p$ = `r signif(lmm2_anova[1,6], 3)`) and *treatment* is not significant ($p$ = `r signif(lmm2_anova[2,6], 3)`), thus showing the opposite result of @eq-rat-lm.

# Exercise 3

```{r}
#| label: load data 3
#| include: false

source("R/exercise_3.R")

```

```{r}
#| label: fig-explore-saw
#| fig-pos: H
#| fig-cap: Exploratory plots of the "Saw" data.
#| fig-asp: .8

# Exploratory Plots
par(mfrow = c(1, 3),
    mai=c(.7, .6, .25, 0.2))
boxplot(ly ~ brand, data = saw_data, col = "lightblue", main = "log(Time) by Brand")
boxplot(ly ~ bark, data = saw_data, col = "lightgreen", main = "log(Time) by Bark")
boxplot(ly ~ species, data = saw_data, col = "lightpink", main = "log(Time) by Species")
par(mfrow = c(1, 1))


```

Model:

$$
\log\left( y_{ijklm} \right) = \mu + \alpha_{i} + \beta_{j} + (\alpha\beta)_{ij} + \gamma_{k} + t_{l} + s_{m} + \epsilon_{ijklm}
$$ {#eq-saw-lmm}

Where:

- $\log\left( y_{ijklm} \right)$ is the log-transformed cutting time for the $i$-th *brand*, $j$-th *bark* status, $k$-th *species*, used by the $l$-th *team*, with the $m$-th specific physical *saw.*

- $\mu$ is the overall average log-transformed cutting time.

- $\alpha_{i}$ is the fixed main effect of the $i$-th saw *brand.*

- $\beta_{j}$ is the fixed main effect of the $j$-th *bark* status (e.g., bark vs. no bark).

- $(\alpha\beta)_{ij}$ is the fixed interaction effect between the $i$-th saw *brand* and the $j$-th *bark* status, representing whether the time saved by debarking depends on which brand of saw is used.

- $\gamma_{k}$ is the fixed main effect of the $k$-th wood *species* (e.g., spruce, pine).

- $t_{l}$ is the random effect of the $l$-th team, assumed to be normally distributed with a mean of zero and a variance of $\sigma_{team}^{2}$ ($t_{l} \sim N\left( 0,\sigma_{team}^{2} \right)$).

- $s_{m}$ is the random effect of the $m$-th physical *saw*, assumed to be normally distributed with a mean of zero and a variance of $\sigma_{saw}^{2}$ ($s_{m} \sim N\left( 0,\sigma_{saw}^{2} \right)$).

- $\epsilon_{ijklm}$ is the residual measurement error, assumed to be normally distributed with a mean of zero and a variance of $\sigma^{2}$ ($\epsilon_{ijklm} \sim N\left( 0,\sigma^{2} \right)$).

 

Fitting this model shows the interaction is not necessary ($p$ = `r signif(drop1_m1[2,6], 3)`) on the interaction when performing a F-test. Thus, the model is reduced to a final model, @eq-saw-final-lmm:

$$
\log\left( y_{ijklm} \right) = \mu + \alpha_{i} + \beta_{j} + \gamma_{k} + t_{l} + s_{m} + \epsilon_{ijklm}
$$ {#eq-saw-final-lmm}

## Findings from the additive model

```{r}
#| label: tbl-fixed-eff-saw-m2
#| tbl-cap: Fixed effects of @eq-saw-final-lmm

sum_m2 <- summary(m2)

sum_m2$coefficients %>% 
  as.data.frame() %>% 
  rename("p-value" = last_col()) %>% 
  kable()

```

As seen @tbl-fixed-eff-saw-m2, *brand* appears to be not significant, however, both *species* and *bark* show statistical significance.

```{r}
#| label: tbl-saw-random
#| tbl-cap: Random effects of @eq-saw-final-lmm

rando <- as.data.frame(VarCorr(m2))

rando %>% 
  select(
    Groups = grp,
    Name = var1,
    Variance = vcov,
    "Std.Dev." = sdcor
  ) %>% 
  kable()


```

@tbl-saw-random shows *saw_id* to have a very small variance, indicating it has very little effect on the final result, e.g., using choice of saw makes little difference.

## Task i)

```{r}
#| label: fig-saw-means
#| fig-cap: Comparison of *saw* brands.
#| fig-asp: .6
#| fig-pos: H


e1t1plot

```

As @fig-saw-means shows, all *brands* perform fairly similar and there can be no statistal results in the difference of cutting time due to brand type, as confidence intervals of all contrast contain 0.

## Task ii)

```{r}
#| label: fig-bark-means
#| fig-cap: Comparison of *bark*.
#| fig-asp: .6
#| fig-pos: H

e3t2plot

```
@fig-bark-means shows that *bark* is significantly different, reducing the cutting time by about 0.15 minutes with fairly high confidence, not contaning 0 ($p$ = <0.0001).

## Task iii)

```{r}
#| label: fig-species-means
#| fig-cap: Comparison of *species*.
#| fig-asp: .6
#| fig-pos: H

par(mai=c(1, 1, .25, 0.5))

e3t3plot

```

According to @fig-species-means, *species* has a statistically significant different effect on cutting times, which is further underlined by the $p$-values shown in @tbl-species-dif.

```{r}
#| label: tbl-species-dif
#| tbl-cap: Effect of *species* on cutting times.

species_df %>% 
  kable()

```

\clearpage

\appendix

# Appendix: Output

```{r}
#| include: false

source("R/exercise_1.R")
```

```{r}
#| echo: true
#| message: true

final_model <- m1

# ranova
ranova(final_model)

# drop1
drop1(final_model, test="F")

# summary
summary(final_model)

```


```{r}
#| include: false

source("R/exercise_2.R")
```

```{r}
#| echo: true

final_model <- m2

# ranova
ranova(final_model)

# drop1
drop1(final_model, test="F")

# summary
summary(final_model)

```

```{r}
#| include: false

source("R/exercise_3.R")
```

```{r}
#| echo: true

final_model <- m2

# ranova
ranova(final_model)

# drop1
drop1(final_model, test="F")

# summary
summary(final_model)

```
\clearpage

# Appendix: Full R code

```{r}
#| echo: false
#| results: asis

files <- list.files("R/", pattern = "\\.R$", full.names = TRUE)

for (f in files) {
  cat("\n\n## ", basename(f), "\n\n", sep = "")
  cat("```{.r}\n")
  cat(readLines(f), sep = "\n")
  cat("\n```\n")
}
```

