---
title: "Assignment 1 - 02429 Analysis of correlated data: Linear mixed models"
subtitle: "Due date: Monday September 28, 2026"
author:
  - name: Lachlan Casey
    email: s263853@student.dtu.dk
  - name: Boglárka Dorottya Kováts
    email: s262008@student.dtu.dk
  - name: Rune Ringgaard Jørgensen
    email: runjor@food.dtu.dk
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{float}
execute: 
  echo: false
bibliography: references.bib
---

# Exercise I. Experimental data on heritability. A components of variance model.

<!-- Objectives: -->

<!-- - Estimate the percentage of conceptions with a confidence interval. -->

<!-- - Report the variability in the percentages due to bull-to-bull variability and within-bull variability (components of variance). And the heritability parameter. -->

```{r}
#| label: load data
#| include: false

source("R/exercise-1.R")

```

Exploring the data, plots are created to visualise the data, shown in @fig-bulls-plots. @fig-boxplot-bulls shows the percentage of conception by bull, displayed in box plot. @fig-observations shows observations of conception for each sample of bull semen.

::: {#fig-bulls-plots layout="[[48,-4,48]]" fig-pos="H"}
```{r}
#| label: fig-boxplot-bulls
#| echo: false
#| fig-cap: Mean conception of bulls. Red dots = means.
#| fig-width: 2.5
#| fig-asp: 1.2

plot_bulls

```

```{r}
#| label: fig-observations
#| echo: false
#| fig-cap: Observations of conception for bulls. Red squares = means.
#| fig-width: 2.5
#| fig-asp: 1.2
#| fig-align: center

plot_observations

```

Conception data visualised by bull.
:::

The objective of the exercise is to describe the variation within the bull population the sample is taken from. Therefore, observations from each bull must be seen as representations of the population, meaning the individual bulls are not important. The six bulls are considered a random sample from a larger population of bulls. Consequently, inference is directed towards the population variability rather than towards the performance of the individual bulls.

@fig-bulls-plots shows a large variance in the samples from bull 4, while others, e.g., bull 3 have a lower variance. Furthermore, bull 5 has a high percentage of conception compared to the rest of the sample.

### Estimation of percentage of conception

Considering the bulls as a random sample of a population, and assuming that observations from each bull are correlated, the percentage of conception can be estimated with a random-effects model with bull as a random factor:

$$
y_{ij}=\mu+b_i+e_{ij}
$$ {#eq-model}

Where:

$y_{ij}$ =

: conception percentage for sample $j$ from bull $i$

$\mu$ =

: overall mean conception percentage

$b_i$ =

: random effect of bull $i$

$e_{ij}$ =

: residual variation

 

Using RStudio software (v. `r base::getRversion()`) with package `'lme4'` (v. `r base::getNamespaceVersion(ns = "lme4")`), a linear mixed model is computed as [@eq-model; @lme4]. The population mean conception percentage was estimated using the intercept from the mixed model:

**`r estimate`**

## Components of variance

The fitted random-effects model allows the total variability in conception percentage to be separated into variability between bulls and variability within bulls. The random bull effect, $b_i$, represents systematic differences between bulls, whereas the residual term, $e_{ij}$, represents variation between samples from the same bull.

<!-- The variance components were extracted from the fitted mixed model. The estimated between-bull variance $\hat{\sigma}_a^2 = `r sigma_bull` $, while the estimated between-bull variance was $\hat{\sigma}_\varepsilon^2 = `r sigma_within` $. Consequently, the total variance can be decomposed as: -->

The variance components were extracted from the fitted mixed model. The estimated between-bull variance $\hat{\sigma}_a^2 =$ `r round(sigma_bull, 1)`, while the estimated between-bull variance was $\hat{\sigma}_\varepsilon^2 =$ `r round(sigma_within, 1)`. Consequently, the total variance can be decomposed as @eq-totalvariance:

$$
\hat{\sigma}_y^2 = \hat{\sigma}_a^2 + \hat{\sigma}_\varepsilon^2
$$ {#eq-totalvariance}

This decomposition shows how much of the overall variability in conception percentages is attributable to differences between bulls and how much is due to variability among semen samples from the same bull. The within-bull variance was larger than the between-bull variance, indicating substantial variation among samples from individual bulls.

## Intraclass correlation and heritability

The intraclass correlation coefficient (ICC) was calculated as @eq-ICC:

$$
\rho = \frac{\sigma_a^2}{\sigma_a^2+\sigma_\varepsilon^2}
$$ {#eq-ICC}

The ICC measures the correlation between two observations originating from the same bull and can also be interpreted as the proportion of total variance attributable to differences between bulls. An ICC of indicates that approximately of the total variation in conception percentage is explained by bull-to-bull differences, while the remaining variation is due to variability among samples from the same bull. Following the lecture notes, the heritability parameter was calculated as @eq-heredity:

$$
h = \frac{4\sigma_a^2}{\sigma_a^2+\sigma_\varepsilon^2} =
\frac{4 × `r round(sigma_bull, 1)`}{`r round(sigma_bull, 1)` + `r round(sigma_within, 1)`}
$$ {#eq-heredity}

The estimated heritability was $h =$ `r round(h, 1)`. A high value indicates that a substantial proportion of the variation in conception percentage can be attributed to differences between bulls, whereas a low value would indicate that most variation occurs within bulls rather than between them.

## Conclusion

A random-effects model was used to estimate the mean conception percentage and quantify the different sources of variability. The estimated mean conception percentage was `r estimate`. The variance decomposition showed that variability existed both between bulls and within bulls. The ICC further demonstrated that observations from the same bull were positively correlated, supporting the use of a mixed model rather than a simple linear model.

# References

::: {#refs}
:::
