---
title: "Assignment 1 - 02429 Analysis of correlated data: Linear mixed models"
subtitle: "Due date: Monday September 28, 2026"
author:
  - name: Lachlan Casey
    email: s263853@student.dtu.dk
  - name: Boglárka Dorottya Kováts
    email: s262008@student.dtu.dk
  - name: Rune Ringgaard Jørgensen
    email: runjor@food.dtu.dk
format:
  pdf:
    include-in-header:
      text: |
        \usepackage{float}
execute: 
  echo: false
---

## Exercise I. Experimental data on heritability. A components of variance model.

<!-- Objectives: -->

<!-- - Estimate the percentage of conceptions with a confidence interval. -->

<!-- - Report the variability in the percentages due to bull-to-bull variability and within-bull variability (components of variance). And the heritability parameter. -->

```{r}
#| label: load data
#| include: false

source("R/exercise-1.R")

```

Exploring the data, plots are created to visualise the data, shown in @fig-bulls-plots. @fig-boxplot-bulls shows the percentage of conception by bull, displayed in box plot. @fig-observations shows observations of conception for each sample of bull semen.

::: {#fig-bulls-plots layout="[[48,-4,48]]" fig-pos="H"}
```{r}
#| label: fig-boxplot-bulls
#| echo: false
#| fig-cap: Mean conception of bulls. Red dots = means.
#| fig-width: 2.5
#| fig-asp: 1.2

plot_bulls

```

```{r}
#| label: fig-observations
#| echo: false
#| fig-cap: Observations of conception for bulls. Red squares = means.
#| fig-width: 2.5
#| fig-asp: 1.2
#| fig-align: center

plot_observations

```

Conception data visualised by bull.
:::

### Estimation of percentage of conception

Considering the bulls as a random sample of a population, and assuming that observations from each bull are correlated, the percentage of conception can be estimated with a random-effects model with bull as a random factor:

$$
y_{ij}=μ+b_i​+e_{ij}​
$$ {#eq-model}

Where:

$y_{ij}$ =

: conception percentage for sample $j$ from bull $i$

$μ$ =

: overall mean conception percentage

$b_i$ =

: random effect of bull $i$

$e_{ij}$ =

: residual varition

 

The estimation of percentage of conception with a 95% confidence interval is:

`r estimate`
